Algebraic Tori in Unitary Groups

### 2014-06-27 Paper.tex
Changes at CNTA

\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\PGL}{PGL}
\DeclareMathOperator{\SL}{SL}

@@

TODO book of involutions 29.20 says this is trivial.
TODO --- Find reference for this ----

@@

We may thus define the group schemes $SO_J$ and $SP$:
\[ SO_J(R) =  \{ g\in (M_{n}(C) \otimes_K R)^\times \mid \tau_C(g^t)Jg = J  \} \]
and
\[ SP(R) =  \{ g\in (M_{n}(C) \otimes_K R)^\times \mid \sigma_C(g^t)Jg = J  \}. \]
We redilly observe that $SU_J$ embedds naturally into both of these groups.
\begin{remark}
The group $SP$ is a symplectic group, its isomorphism class does not depend on $J$.
\end{remark}

\section{\'Etale Algebras  associated to Tori in Algebra Groups}

@@

\subsection{Tori in Algebraic Groups}

@@

\section{Forms preserved by $T_{E,\sigma}$}

\subsection{Shape of the Forms}

@@

\subsection{Invariants of Hermitian Forms}

@@

\subsection{Existence of Forms}

@@

\begin{theorem}
Let $K$ be a number field and fix $A$ a quadratic extension of $K$ (potentially split).
Given $H$ a Hermitian form and $(E,\sigma)$ an \'etale algebra with involution (of the second kind) over $A$. Then there exists $\lambda \in E^\sigma$ with:
\[ H \simeq H_{E,\sigma,\lambda} \]
if and only if there exists $\lambda_\fp$ for all localizations $\fp$ of $K$ (both archimedian and non-archimedian).
\end{theorem}
\begin{proof}
\TODO - complete sketch (also check that this is true).
\TODO - need to distinguish primes of $K$, $A$ and $E$.

Write $E^\sigma = \prod_{i=1}^n E^\sigma_i$.
We have the following diagram:
\[
\xymatrix{
(E^\sigma)^\times/N_{E/E^\sigma}(E^\times) \ar[r]^{(\cdot,z)} \ar[d]& H^2(E^\sigma, \pm 1) \ar[r] \ar[d]& \underset{E_{i,\fp} \text{ not split}}\oplus H^2(E^\sigma_{i,\fp},\pm 1) \ar[d]\\
K^\times/N_{A/K}(A^\times) \ar[r]^{(\cdot,\delta_{A/K})} & H^2(K, \pm 1) \ar[r] & \underset{A_\fp \text{ not split}}\oplus H^2(K_\fp,\pm 1)
}
\]
where the compositions of the horizontal maps are injective.
We must show that an element in $K^\times/N_{A/K}(A^\times) $ is in the image of the left verticle map, provided its image is in the image of the right hand verticle map.
We must do this in spite of the fact that the horizontal maps are not surjective.

@@

\begin{theorem}
\TODO - summarize a pure global existance statement.
\end{theorem}

\begin{remark}
\TODO - Can phrase this as: The reflex algebra splits the even clifford algebra + signature conditions.
Is there a meaningfull interpretation of the clifford algebra in this context?
Can we interpret it as telling us about a quaternioninc symplectic group embedding of the unitary group?
\end{remark}

\subsection{Classification of Tori in Unitary Groups}

@@

Let $(E,M_n,\tilde{\tau})$ be the triple consisting of $E$, the matrix algebra of degree $n$ over $E$, and an involution $\tilde{\tau}$ of disciminant $\delta_{\tau}$ so that $U_{E,A,\tau}$ is the unitary group for the $E$-Hermitian space of discriminant $\delta_{\tau}$.

The rational conjugacty classes of maximal tori in $U_{E,A,\tau}$ are naturally identified with a subset of those of $U_{E,M_n,\tilde{\tau}}$.
The tori in $U_{E,M_n,\tilde{\tau}}$ are assicated to triples $(F,\sigma,\lambda)$, the rational conjugacy classes of tori in $U_{E,A,\tau}$ are precisely the subset for which $(F,\sigma)\hookrightarrow (A,\tau)$, equivalently, the subset for which $F$ splits $A$.

The tori associated to $(F,\sigma,\lambda)$ in $U_{E,A,\tau}$ is the same as that in $U_{E,M_n,\tilde{\tau}}$ in particular they are $T_{F,\sigma}$, fixing an embedding $\rho : F \rightarrow A$, and an involution $\hat\tau$ of disciminant $1$ the conjugacy classes correspond to the image of $T_{F,\sigma}$ in $U_{E,A,\tau} \subset A$ where $U_{E,A,\tau}$ is defined comes from $\tau(x) = \rho(\lambda^{-1})\hat\tau(x)\rho(\lambda)$.

@@

Local existance -> global existance is easy, as $\lambda \in F$ places $\lambda \in A$ so as to allow us to define the involution, which has the correct local invariants and thus gives the correct unitary group.

@@

\section{Properties of $T_{E,\sigma}$}

\subsection{Anisotropic}

\subsection{Unramified}

\subsection{``Relatively'' Unramified}

\section{The Building of the Unitary Group}

\subsection{The reduction of $T_{E,\sigma}$}

@@

\begin{theorem}
Suppose $H$ is a Hermitian space
\end{theorem}

>>>>

\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\PGL}{PGL}
\DeclareMathOperator{\SL}{SL}
\DeclareMathOperator{\Sp}{Sp}
\DeclareMathOperator{\SO}{SO}
\DeclareMathOperator{\U}{U}
\DeclareMathOperator{\SU}{SU}

@@

TODO book of involutions 29.20 says this is trivial.

@@

\section{Classifying Tori in Classical Groups}
%\section{\'Etale Algebras associated to Tori in Algebra Groups}

\TODO{Edit section to align with the following goal}

In this section we describe maximal tori in classical groups in terms of \'etale algebras with involutions.

@@

\subsection{Tori in algebraic groups}

@@

\subsection{Intrinsic properties of these tori}

\subsubsection{Anisotropic}

\subsubsection{Unramified}

\subsubsection{``Relatively'' unramified}

\section{Classical groups containing a fixed torus}

For each torus constructed in the previous section, we describe into which classical groups it embeds.  We also describe the conjugacy classes of embeddings, both stably and rationally.

\subsection{Constructing bilinear forms from tori}

@@

\subsection{Invariants of the bilinear forms}

@@

\subsection{Embedding in different forms}

@@

\begin{theorem}
Let $K$ be a number field and fix $A$ a quadratic extension of $K$ (potentially split).
Given $H$ a Hermitian form and $(E,\sigma)$ an \'etale algebra with involution (of the second kind) over $A$. Then there exists $\lambda \in E^\sigma$ with:
\[ H \simeq H_{E,\sigma,\lambda} \]
if and only if there exists $\lambda_\fp$ for all localizations $\fp$ of $K$ (both archimedean and non-archimedean).
\end{theorem}
\begin{proof}
\TODO - complete sketch (also check that this is true).
\TODO - need to distinguish primes of $K$, $A$ and $E$.

Write $E^\sigma = \prod_{i=1}^n E^\sigma_i$.
We have the following diagram:
\[
\xymatrix{
(E^\sigma)^\times/N_{E/E^\sigma}(E^\times) \ar[r]^{(\cdot,z)} \ar[d]& H^2(E^\sigma, \pm 1) \ar[r] \ar[d]& \underset{E_{i,\fp} \text{ not split}}\oplus H^2(E^\sigma_{i,\fp},\pm 1) \ar[d]\\
K^\times/N_{A/K}(A^\times) \ar[r]^{(\cdot,\delta_{A/K})} & H^2(K, \pm 1) \ar[r] & \underset{A_\fp \text{ not split}}\oplus H^2(K_\fp,\pm 1)
}
\]
where the compositions of the horizontal maps are injective.
We must show that an element in $K^\times/N_{A/K}(A^\times) $ is in the image of the left vertical map, provided its image is in the image of the right hand vertical map.
We must do this in spite of the fact that the horizontal maps are not surjective.

@@

\begin{theorem}
\TODO{summarize a pure global existence statement.}
\end{theorem}

\begin{remark}
\TODO{Can phrase this as: The reflex algebra splits the even clifford algebra + signature conditions.
Is there a meaningful interpretation of the clifford algebra in this context?
Can we interpret it as telling us about a quaternionic symplectic group embedding of the unitary group?}
\end{remark}

\subsection{Rational Conjugacy of Tori}

@@

Let $(E,M_n,\tilde{\tau})$ be the triple consisting of $E$, the matrix algebra of degree $n$ over $E$, and an involution $\tilde{\tau}$ of discriminant $\delta_{\tau}$ so that $U_{E,A,\tau}$ is the unitary group for the $E$-Hermitian space of discriminant $\delta_{\tau}$.

The rational conjugacy classes of maximal tori in $U_{E,A,\tau}$ are naturally identified with a subset of those of $U_{E,M_n,\tilde{\tau}}$.
The tori in $U_{E,M_n,\tilde{\tau}}$ are associated to triples $(F,\sigma,\lambda)$, the rational conjugacy classes of tori in $U_{E,A,\tau}$ are precisely the subset for which $(F,\sigma)\hookrightarrow (A,\tau)$, equivalently, the subset for which $F$ splits $A$.

The tori associated to $(F,\sigma,\lambda)$ in $U_{E,A,\tau}$ is the same as that in $U_{E,M_n,\tilde{\tau}}$ in particular they are $T_{F,\sigma}$, fixing an embedding $\rho : F \rightarrow A$, and an involution $\hat\tau$ of discriminant $1$ the conjugacy classes correspond to the image of $T_{F,\sigma}$ in $U_{E,A,\tau} \subset A$ where $U_{E,A,\tau}$ is defined comes from $\tau(x) = \rho(\lambda^{-1})\hat\tau(x)\rho(\lambda)$.

@@

Local existence $\to$ global existence is easy, as $\lambda \in F$ places $\lambda \in A$ so as to allow us to define the involution, which has the correct local invariants and thus gives the correct unitary group.

@@

\begin{itemize}
\item When does $\U_n$ embed into $\SO_{2n}$, $\SO_{2n+1}$ and $\Sp_{2n}$.
\item Which tori in $\SO_m / \Sp_{2n}$ are in the images these embeddings.
\item What are the rational conjugacy classes of these embeddings, and how do the rational conjugacy classes of tori break up?
\item Do different inner forms embed into the same place?  What is the intersection?

if K is non-split then the Hasse invariant of the resulting orthogonal group depends on isomorphism class of Hermitian space and thus they don't embed into the same group.

If K is split, then the resulting orthogonal space is determined entirely by the choice of quaternion algebra (split/ramified) for the construction of the unitary group. We obtain different orthogonal groups in the 2 cases.
\TODO{check how many orthogonal group isomorphism classes in this case}

Thus there is no intersection.

\item Can we understand tori in adjoint forms (earlier section).  For $\SO$ and $\Sp$ you quotient by $\pm 1$ while for $\U_n$ you quotient by $\U_1$.  Describe these as subtori instead of quotients.
\end{itemize}

We may thus define the group schemes $SO_J$ and $SP$:
\[ SO_J(R) =  \{ g\in (M_{n}(C) \otimes_K R)^\times \mid \tau_C(g^t)Jg = J  \} \]
and
\[ SP(R) =  \{ g\in (M_{n}(C) \otimes_K R)^\times \mid \sigma_C(g^t)Jg = J  \}. \]
We readilly observe that $SU_J$ embeds naturally into both of these groups.
\begin{remark}
The group $SP$ is a symplectic group, its isomorphism class does not depend on $J$.
\end{remark}

@@
